---
title: "IDS investigation worksheet"
author: "by Team-Name: User1, User2, User3, User4 & User5"
date: "`r Sys.Date()`"
output: html_document
---

**Note:** You can use this file as you 'working document' where you can try out various investigation ideas and keep notes about your findings. How you use and structure this file is up to you. It is recommended that you keep notes about what you are investigating and what you find as this will make the process of creating your presentation and report easier. Please note that you _do not_ need to submit this file as part of your group project.



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r load-lib, message = FALSE}
library(tidyverse)
library(readr)
library(dplyr)
library(janitor)

```



create_regions creates a csv file called file_name, filtering nvidia by reg. 
create_csv creates dataframes for each region in nvidia (I ran create_csv(regions, file_names) in console). 
region_dfs is a list of all dataframes by region. 

```{r load-data}
olympics <- read_csv("~/Downloads/olympics.csv")
```

``` {r}

olympics_medals <-
  olympics %>%
  drop_na()

#finding win percentage overall
olympics %>% 
  group_by(medal) %>% 
  summarise(count = n()) %>%
  mutate(win_pct = count/sum(count) * 100 )

# athletes by team
olympics %>%
  group_by(team) %>% 
  summarise(count = n()) %>%
  mutate(team_pct = count/sum(count) * 100 ) %>%
  arrange(desc(team_pct))

# medals by team
olympics_medals %>%
  group_by(team) %>% 
  summarise(count = n()) %>%
  mutate(medals_pct = count/sum(count) * 100 ) %>%
  arrange(desc(medals_pct))

# teams in sports
olympics_medals %>%
  count(team, sport) %>%
  arrange(desc(n))

```

